ICM testing

### 2023-03-15 Scripts/ASR_stocks.Rmd
Fixed fecundity offset
Had to wrap my head around making sure the mx vector was in the correct year... had to account for offset of recruits showing up in the data.  Was painful... think I got it now.

The data is now avilable to test this on at least 81 stocks.  All of the NE Atlantic stocks (ICES) have been through the QA/QC process and we are happy with those data (Sept 1, 2022).  Let's see if we can get this to work on all the stocks with sufficient data...

@@

# Replace 0's in Num.tot with NA so the rest of this works as Freya designed it to...
ASR$Num.tot[ASR$Num.tot == 0] <- NA
# we want:
# Year, Num, WA, Catch, AM, NM, StockID, Management, Area, Order, Family, Genus, Species
ASRdat <- ASR[,c(grep(x=names(ASR), "Num"),
                 grep(x=names(ASR), "WA"),
                 grep(x=names(ASR), "Catch"),
                 grep(x=names(ASR), "AM"),
                 grep(x=names(ASR), "NM"))]
# Making all the data numberic that should be numeric
ASRdat <- apply(X = ASRdat, 2, as.numeric)
# Getting the species info back
ASRsp <- ASR[, which(!1:length(names(ASR)) %in% grep(x=names(ASR), ".", fixed=T))]
# And binding it all back together
ASR_trim <- cbind(ASRsp, ASRdat)

@@

# summary plots for ALL stocks
for (i in unique(ASR_long$Stock)) {
  ASR_sub <- ASR_long %>%
    dplyr::select(Year, Stock, type, age, value) %>%
    dplyr::filter(Stock==i)
  maxage <- max(as.numeric(ASR_sub$age[!is.na(ASR_sub$value)]))
  maxYear <- max(ASR_sub$Year[!is.na(ASR_sub$value)])
  minYear <- min(ASR_sub$Year[!is.na(ASR_sub$value)])
  forplot <- ASR_sub[as.numeric(ASR_sub$age) < (maxage+1) & ASR_sub$Year %in% minYear:maxYear,]
  p1 <- ggplot() + geom_line(data=forplot, aes(Year, value)) + facet_grid(type~as.numeric(age), scales="free_y") +
        geom_vline(data=forplot[is.na(forplot$value),], aes(xintercept=Year), colour="red") + facet_grid(type~as.numeric(age), scales="free_y") +
        theme_bw() + 
        ggtitle(i) +
    guides(x =  guide_axis(angle = 90)) 
  print(p1)

@@

  rem <- data %>% dplyr::group_by(Year) %>% dplyr::summarize(rem=sum(Catch,na.rm=T)) %>% dplyr::pull(rem)

  missing_rem<- NULL
  if(any(is.na(rem))) {
    missing_rem <- unique(data$Year)[which(is.na(rem))]
    rem[is.na(rem)] <- median(rem, na.rm=T)
  }

  #rowSums(removals[,-which(names(removals) %in% c("Year", "Stock", "type"))], na.rm=T)
  years <- data %>% pull(Year) %>% unique() %>% sort()
  N.end <- sum(data[data$Year==max(years),]$Num)
  #N.end <- rowSums(abund[nrow(abund),-which(names(abund) %in% c("Year", "Stock", "type"))], na.rm=T)
  #vpa.abund <- rowSums(abund[,-which(names(abund) %in% c("Year", "Stock", "type"))], na.rm=T)
  vpa.abund <- data %>% dplyr::group_by(Year) %>% dplyr::summarize(vpa=sum(Num)) %>% pull(vpa)

  # The real mx matrix, recruits produced per individual in each age class... Not perfect as I need to offset recruits/ssb, but close enough for the moment..
  #minage
  minage <- min(as.numeric(data$age))
  #maxage
  maxage <- max(as.numeric(data$age))
  #recruits
  annual <- data.frame(Year=data$Year[data$age==minage], recruits=data$Num[data$age==minage])
  #ssn
  data$ssn <- data$Num * data$AM
  #ssb
  data$ssb <- data$ssn * data$WA
  #tot.ssb
  annual <- data %>% group_by(Year) %>% summarize(tot.ssb = sum(ssb)) %>% left_join(annual)
  #r.p.ssb
  annual$r.p.ssb <- annual$recruits/annual$tot.ssb
  # recs.per.age
  data <- left_join(data, annual)
  data$recs.per.age <- data$ssb*data$r.p.ssb
  # mx
  data$mx <- data$recs.per.age/data$ssn/2 # Moms only! Dividing by around 8 really nails it for NS cod for whatever reason :-)
  data$mx[is.nan(data$mx)] <- 0 # if we don't have any spawners in an age class in a year their fecundity is 0
  # Something is wrong with the decline rate method, but the exponential and logistic are working pretty... pretty... pretty good...

  age.mat <- data %>% dplyr::select("Year", "age", "AM") %>% pivot_wider(names_from=age, values_from = AM) %>% dplyr::select(-Year)
  prop.nat.mort <- data %>% dplyr::select("Year", "age", "prop.nat.mort") %>% pivot_wider(names_from=age, values_from = prop.nat.mort) %>% dplyr::select(-Year)
  weight.age <- data %>% dplyr::select("Year", "age", "WA") %>% pivot_wider(names_from=age, values_from = WA) %>% dplyr::select(-Year)
  mx <- data %>% dplyr::select("Year", "age", "mx") %>% pivot_wider(names_from=age, values_from = mx) %>% dplyr::select(-Year) %>% as.data.frame()

  tst <- icm.sim(years = years,
                 mat.age = age.mat,
                 nm = prop.nat.mort,
                 w.age = weight.age,
                 ages = minage:maxage,
                 rems = rem,
                 fecund = mx,
                 N.end = N.end,
                 pop.model = 'exponential',
                 n.sims = 500,
                 sd.mat = 0.5,
                 sd.nm = 0.5,
                 sd.wt = 0.5,
                 sd.fecund = 0.5)

  #Combine the data
  did.it.work <- data.frame(abund = c(vpa.abund,tst$Pop$abund),years = c(years,tst$Pop$years),sim = c(rep('VPA',length(years)),tst$Pop$sim))
  # Here get the Upper and lower 50% quantiles to make a functional boxplot
  quants <- did.it.work %>% dplyr::group_by(years) %>% dplyr::summarise(L.50 = quantile(abund,probs=c(0.25)),
                                                                        med = median(abund),
                                                                        U.50 = quantile(abund,probs=c(0.75)))

@@

  # All the results in a plot
  p2 <- ggplot() +
          geom_line(data=did.it.work, aes(x=years,y=abund,color=sim)) +
          xlab("") +
          ylab("Abundance (1000s)") +
          geom_line(data=did.it.work %>% dplyr::filter(sim == "VPA"),aes(x=years,y=abund),color='black',size=2) +
          #scale_y_continuous(breaks = seq(0,3e6,by=5e5)) + scale_x_continuous(breaks = seq(1960,2025,by=5)) +
          theme_few() + theme(legend.position = 'none') + scale_color_viridis_d(end = 0.75) +
          ggtitle(i) +
          geom_text(data=did.it.work[did.it.work$missing=="missing",], aes(x=years, y=abund, label="?"))
  # Same thing but functional boxplots
  p3 <- ggplot() +
          geom_line(data=quants, aes(x=years,y= med)) +
          geom_ribbon(data=quants, aes(x=years,ymax=U.50,ymin = L.50),alpha=0.5,fill='blue',color='blue') +
          geom_line(data=did.it.work %>% dplyr::filter(sim == "VPA"),aes(x=years,y=abund),color='black',size=2) +
          xlab("") + ylab("Abundance (1000s)") +
          #scale_y_continuous(breaks = seq(0,3e6,by=5e5)) + scale_x_continuous(breaks = seq(1960,2025,by=5)) +
          theme_few() + theme(legend.position = 'none') +
          ggtitle(i)+
          geom_text(data=did.it.work[did.it.work$missing=="missing",], aes(x=years, y=abund, label="?"))
  # Plotting how the estimate of r changes over time (which you can do when we have time varying inputs (e.g. Wgt, M, or maturity at age)
  p4 <- ggplot(tst$r) +
          geom_line(aes(x=years,y=r,group=n.sims,color=n.sims)) +
          xlab("") +
          ylab("Estimated growth rate (r)") +
          #scale_y_continuous(breaks = seq(0,1.5,by=0.1)) +
          #scale_x_continuous(breaks = seq(1960,2025,by=5)) +
          theme_few() + theme(legend.position = 'none') +
          scale_color_viridis_c(end = 0.75) +
          ggtitle(i)

  print(p2)
  print(p3)
  print(p4)

>>>>

The data is now available to test this on at least 81 stocks.  All of the NE Atlantic stocks (ICES) have been through the QA/QC process and we are happy with those data (Sept 1, 2022).  Let's see if we can get this to work on all the stocks with sufficient data...

@@

# Replace 0's in Num.tot with NA so the rest of this works as Freya designed it to...
ASR$Num.tot[ASR$Num.tot == 0] <- NA
# we want:
# Year, Num, WA, Catch, AM, NM, StockID, Management, Area, Order, Family, Genus, Species
ASRdat <- ASR[,c(grep(x=names(ASR), "Num"),
                 grep(x=names(ASR), "WA"),
                 grep(x=names(ASR), "Catch"),
                 grep(x=names(ASR), "AM"),
                 grep(x=names(ASR), "NM"))]
# Making all the data numeric that should be numeric
ASRdat <- apply(X = ASRdat, 2, as.numeric)
# Getting the species info back
ASRsp <- ASR[, which(!1:length(names(ASR)) %in% grep(x=names(ASR), ".", fixed=T))]
# And binding it all back together
ASR_trim <- cbind(ASRsp, ASRdat)

@@

# summary plots for ALL stocks
for (i in unique(ASR_long$Stock)) {
  ASR_sub <- ASR_long %>%
    dplyr::select(Year, Stock, type, age, value) %>%
    dplyr::filter(Stock==i)
  maxage <- max(as.numeric(ASR_sub$age[!is.na(ASR_sub$value)]),na.rm=T)
  maxYear <- max(ASR_sub$Year[!is.na(ASR_sub$value)],na.rm=T)
  minYear <- min(ASR_sub$Year[!is.na(ASR_sub$value)],na.rm=T)
  forplot <- ASR_sub[as.numeric(ASR_sub$age) < (maxage+1) & ASR_sub$Year %in% minYear:maxYear,]
  forplot <- forplot[!is.na(forplot$value),] # remove the NAs, hopefully this works for all the stocks...
  p1 <- ggplot() + geom_line(data=forplot, aes(Year, value)) + facet_grid(type~as.numeric(age), scales="free_y") +
        geom_vline(data=forplot[is.na(forplot$value),], aes(xintercept=Year), colour="red") + facet_grid(type~as.numeric(age), scales="free_y") +
        theme_bw() + 
        ggtitle(i) +
    guides(x =  guide_axis(angle = 90)) 
  print(p1)

@@

  rem <- data %>% dplyr::group_by(Year,.drop=F) %>% dplyr::summarize(rem=sum(Catch,na.rm=T)) #%>% dplyr::pull(rem)

  missing_rem<- NULL
  if(any(is.na(rem$rem))) 
  {
    missing_rem <- unique(data$Year)[which(is.na(rem$rem))]
    rem$rem[is.na(rem$rem)] <- median(rem$rem, na.rm=T)
  }

  #rowSums(removals[,-which(names(removals) %in% c("Year", "Stock", "type"))], na.rm=T)
  years <- data %>% dplyr::pull(Year) %>% unique() %>% sort()
  N.end <- sum(data[data$Year==max(years),]$Num)
  #N.end <- rowSums(abund[nrow(abund),-which(names(abund) %in% c("Year", "Stock", "type"))], na.rm=T)
  #vpa.abund <- rowSums(abund[,-which(names(abund) %in% c("Year", "Stock", "type"))], na.rm=T)
  vpa.abund <- data %>% dplyr::group_by(Year) %>% dplyr::summarize(vpa=sum(Num)) %>% pull(vpa)

  # The real mx matrix, recruits produced per individual in each age class... Not perfect as I need to offset recruits/ssb, but close enough for the moment..
  #minage
  minage <- min(as.numeric(data$age))
  #maxage
  maxage <- max(as.numeric(data$age))
  #recruits
  annual <- data.frame(Year=data$Year[data$age==minage], recruits=data$Num[data$age==minage])
  #ssn
  data$ssn <- data$Num * data$AM
  #ssb
  data$ssb <- data$ssn * data$WA #I need to figure out how to line this up with the number of recruits for the right years in the data object.... got it right for the overall r.p.ssb below.
  #tot.ssb
  annual <- data %>% group_by(Year) %>% summarize(tot.ssb = sum(ssb)) %>% left_join(annual)
  #r.p.ssb This needs to be offset by the age of recruits
  annual$r.p.ssb <- c(rep(NA,minage),annual$recruits[(minage+1):nrow(annual)]/annual$tot.ssb[1:(nrow(annual)-minage)])
  # recs.per.age
  data <- left_join(data, annual)
  # So I need to make an offset SSB in the 'data' object to line up with the correct r.p.ssb field.  This is probably gonna suck...
  tmp <- NULL
  for(j in 1:length(years))
  {
    tst <- data %>% dplyr::filter(Year == years[j]) %>% dplyr::select(ssb,ssn,Year,age)
    tst$Year <- tst$Year + minage
    names(tst) <- c("ssb.offset","ssn.offset","Year","age")
    tmp[[j]] <- tst
  }
  # Unpack the list
  ssb.off <- do.call('rbind',tmp)
  # And merge it with the data object
  data <- left_join(data,ssb.off,by=c("Year","age"))
  data$recs.per.age <- data$ssb.offset*data$r.p.ssb
  # mx
  # Some of the stocks are just the males and females, so these are already half the population.  Probably can just not include the male stocks in the end.
  # This works because 'males' is also in 'females'
  if(grepl("males",i)) data$mx <- data$recs.per.age/data$ssn.offset 
  if(!grepl("males",i))data$mx <- data$recs.per.age/data$ssn.offset/2 # Moms only! 
  data$mx[is.nan(data$mx)] <- 0 # if we don't have any spawners in an age class in a year their fecundity is 0
  # Easier to remove the years where we don't have recs.per.age...
  #data <- data %>% dplyr::filter(Year %in% years[(minage+1):length(years)])
  # Something is wrong with the decline rate method, but the exponential and logistic are working pretty... pretty... pretty good...

  age.mat <- data %>% dplyr::select("Year", "age", "AM") %>% pivot_wider(names_from=age, values_from = AM) %>% dplyr::select(-Year)
  prop.nat.mort <- data %>% dplyr::select("Year", "age", "prop.nat.mort") %>% pivot_wider(names_from=age, values_from = prop.nat.mort) %>% dplyr::select(-Year) %>% as.data.frame
  weight.age <- data %>% dplyr::select("Year", "age", "WA") %>% pivot_wider(names_from=age, values_from = WA) %>% dplyr::select(-Year)
  mx <- data %>% dplyr::select("Year", "age", "mx") %>% pivot_wider(names_from=age, values_from = mx) %>% dplyr::select(-Year) %>% as.data.frame()
  # For the years of mx without SSB info to inform them
  if(minage >0 ) 
  {
    mx.fill <- as.data.frame(matrix(rep(colMeans(mx,na.rm=T),minage),nrow=minage,byrow=T),colnames = names(mx))
    names(mx) <- names(mx.fill)
   # mx <- rbind(mx[(minage+1):nrow(mx),],mx.fill) # or is it
    mx <- rbind(mx.fill,mx[(minage+1):nrow(mx),]) # I think this is right!
  }
  source("D:/Github/ICM/Scripts/functions/backwards_sim.R")

  tst <- icm.sim(years,
                 mat.age = age.mat,
                 nm = prop.nat.mort,
                 w.age = weight.age,
                 ages = minage:maxage,
                 rems = rem$rem,
                 fecund = mx,
                 N.end = N.end,
                 pop.model = 'exponential',
                 n.sims = 5,
                 sd.mat = 0.5,
                 sd.nm = 0.5,
                 sd.wt = 0.5,
                 sd.fecund = 0.5)

  #Combine the data
  did.it.work <- data.frame(abund = c(vpa.abund,tst$Pop$abund),years = c(years,tst$Pop$years),sim = c(rep('VPA',length(years)),tst$Pop$sim))
  # Here get the Upper and lower 50% quantiles to make a functional boxplot
  quants <- did.it.work %>% dplyr::filter(sim != "VPA") %>% dplyr::group_by(years) %>% dplyr::summarise(L.50 = quantile(abund,probs=c(0.25)),
                                                                        med = median(abund),
                                                                        U.50 = quantile(abund,probs=c(0.75)))

@@

  # All the results in a plot
  p2 <- ggplot() +
          geom_line(data=did.it.work, aes(x=years,y=abund,color=sim)) +
          xlab("") +
          ylab("Abundance (1000s)") +
          geom_line(data=did.it.work %>% dplyr::filter(sim == "VPA"),aes(x=years,y=abund),color='black',size=2) +
          #scale_y_continuous(breaks = seq(0,3e6,by=5e5)) + scale_x_continuous(breaks = seq(1960,2025,by=5)) +
          theme_few() + theme(legend.position = 'none') + scale_color_viridis_d(end = 0.75) +
          ggtitle(i) +
          geom_text(data=did.it.work[did.it.work$missing=="missing",], aes(x=years, y=abund, label="?"))
  # Same thing but functional boxplots
  p3 <- ggplot() +
          geom_line(data=quants, aes(x=years,y= med)) +
          geom_ribbon(data=quants, aes(x=years,ymax=U.50,ymin = L.50),alpha=0.5,fill='blue',color='blue') +
          geom_line(data=did.it.work %>% dplyr::filter(sim == "VPA"),aes(x=years,y=abund),color='black',size=2) +
          xlab("") + ylab("Abundance (1000s)") +
          #scale_y_continuous(breaks = seq(0,3e6,by=5e5)) + scale_x_continuous(breaks = seq(1960,2025,by=5)) +
          theme_few() + theme(legend.position = 'none') + #geom_vline()
          ggtitle(i)+
          geom_text(data=did.it.work[did.it.work$missing=="missing",], aes(x=years, y=abund, label="?"))
  # Plotting how the estimate of r changes over time (which you can do when we have time varying inputs (e.g. Wgt, M, or maturity at age)
  p4 <- ggplot(tst$r) +
          geom_line(aes(x=years,y=r,group=n.sims,color=n.sims)) +
          xlab("") +
          ylab("Lotka r") +
          #scale_y_continuous(breaks = seq(0,1.5,by=0.1)) +
          #scale_x_continuous(breaks = seq(1960,2025,by=5)) +
          theme_few() + theme(legend.position = 'none') +
          scale_color_viridis_c(end = 0.75) +
          ggtitle(i)

  #print(p2)
  #print(p3)
  #print(p4)
